---
title: "Introduction to depositWatch"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to depositWatch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

depositWatch is a package to detect Unusual Behaviour in customer deposit accounts Using Rolling Statistical Methods. Traditional Anti-Money Laundering (AML) systems often rely on static thresholds, rule-based alerts, or predefined risk models, which may fail to detect evolving or context-specific suspicious behaviours. depositWatch provides a more adaptive, data-streaming approach to anomaly detection.

### Example
A customer with a typical deposit pattern of $500/month over 2–3 transactions suddenly deposit $50,000 across five transactions. Although each transaction is under the typical reporting threshold, the overall pattern deviates significantly from their norm. Using a rolling Z-score and frequency analysis, such behaviour can be flagged for further review.
This method is dynamic, customer-specific, and scalable, making it suitable for real-world banking applications


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(depositWatch)
```
### Objectives
•	Detect unusual customer deposit behaviour dynamically
•	Allow flexible, user-defined thresholds for anomaly detection
•	Provide clear output for reporting, visualization, and further analysis
•	Maintain scalability and efficiency for large datasets


### Data
The data should be a deposit transactions dataframe having there columns, customer_id, date and amount. 

```{r load-data, echo=TRUE}
data("transactions", package = "depositWatch")
head(transactions)
```

### Functions
The package has four functions:

1. new_deposit_dataset(): This is an S3 constructer that takes a deposit dataset as input and converts it into a "deposit_dataset" class

```{r Input}
#ds = new_deposit_dataset(transactions)
```


2. detect_anomalies(): It takes a deposit_dataset object as input, which contains transaction-level data, and computes rolling statistics such as rolling mean and rolling standard deviation for each customer. The method flags anomalies using two criteria: the first is when a daily total deposit deviates significantly from the rolling mean, exceeding a user-defined Z-score threshold, and the second is when the number of transactions within a day exceeds a frequency threshold, even if the total amount is within normal limits. If both conditions are met simultaneously, the record is flagged as “Both.” The output is a new deposit_dataset object containing a summary of anomalies per customer, including the total count of flagged days and the types of flags encountered.

```{r Results}

```

